### 2021-04-02 assertions/doc/thesis-bernhardt/chapters/4-offline_design_tezos.tex
Type checker

\section{Extensions to Michelson}
The way Michelson is interpreted is purely functional; it takes the current stack and an operation and builds a return stack from the initial one, without causing any side effects \cite{tezos_docs}. The recursive Michelson interpreter is defined as as a list of rules comprising of all possible inputs, i.e. program and stack types, and the respective output stack type of the computation if the rules applies. Each rule is of the following form: 
\begin{lstlisting}[caption=Rules form in the Michelson interpreter \cite{tezos_docs}, language=, numbers=none, label=lst:rules]
> (syntax pattern) / (initial stack pattern)  =>  (result stack pattern)
    iff (conditions)
    where (recursions)
    and (more recursions)
\end{lstlisting}
For each valid program and initial stack exactly one rule matches (given that any extra conditions over values on the stack, stated after the \texttt{iff} keyword are true) \cite{tezos_docs}. If the result depends on the results of other program interpretations (given after the keywords \texttt{iff, where, and} in rule form), the rule only applies if these partial results also match the respective intermediate result stack patterns.\\
In addition, there exists a typing rule for each syntax construct that restricts the valid input stacks. The typing rules use the meta variables \texttt{'a} for type and \texttt{'A} for stack type variables in order to express consistency within the program (not polymorphism). These rules are given in the following form:
\begin{lstlisting}[caption=Form of typing rules \cite{tezos_docs}, language=, numbers=none, label=lst:type_rules]
(syntax pattern)
:: (type of stack before) -> (type of stack after) [rule-name]
   iff (premises)
\end{lstlisting}
The type notations, as well as syntax and stack patterns are listed in \cite{tezos_docs}.

From the given formulas and assertion examples so far, one can derive a set of necessary instructions that need be be present in the target language. The following subsections list the instructions which have to be added to Michelson's instruction set, specify their selection and typing rules and describes how they can be implemented in the interpreter. These do not necessarily include high level instructions, like \texttt{sqrt} from \eqref{eq:prime}, that can be reproduced by using lower level instructions. For these purposes, a later iteration of the syntax will provide the feature of user-defined functions.

\subsection{Random}
A vital instruction that cannot be expressed with Michelson's current instruction set (state of protocol 008 Edo \cite{tezos_docs}) is the generation of a random value of certain types. Since smart contracts need to be deterministic, s.t. the network can reach a consensus about the state of the ledger \cite{chatterjee_probabilistic_2019}, the prevalent workarounds for generating pseudo-random numbers cannot be used for assertion checking. Distributed assertion checking explicitly requires as many validators as possible to generate a unique value, whereas common schemes, like oracles or using block attributes as seeds \cite{chatterjee_probabilistic_2019}, will provide the same values for all validators.\\
\lstref{lst:rand_type} specifies the selection and typing rule of the instruction \texttt{rand}, which consumes an offset and a positive range from the stack and pushes a randomly generated integer to the top. This instruction, of course, can also be provided for other data types, such as \texttt{nat}, \texttt{string} or \texttt{mutez}.
\lstset{upquote=true}
\begin{lstlisting}[caption=Selection and typing rules of the integer \texttt{rand} instruction, language=, numbers=none, label=lst:rand_type]
:: int : nat: 'A   ->  int : 'A

> RAND/ offset : range : S  => int : S
\end{lstlisting}

Within the Michelson interpreter, the selection of rules is implemented as a huge match case \cite{tezos_repo}. If the rule for \texttt{rand} applies, the respective computation in OCaml could look as drafted in \lstref{rand_impl}.
\begin{lstlisting}[caption=Simplified evaluation of \texttt{rand} in the Michelson interpreter, language=, label=lst:rand_impl]
match (instruction, stack) with
...
| (Rand (offset, (range, rest))) -> 
   Random.self_init;
   let rand_int = offset + Random.int(range) in
   return (rand_int, rest)  (* Resulting stack *)
\end{lstlisting}

\subsection{Nth}
An exception for not adding higher level instructions may be the more fundamental operation \texttt{nth} for accessing list elements, as it is used (sometimes more than once, ref. \eqref{eq:sorted_v2}) in many use cases. Adding an own predefined primitive for this reduces the origination costs of the assertion contract, given that its equivalent in the current instruction set requires the compiler to generate a workaround using iteration, for instance a loop or list iterator. As a first approximation, appendix \ref{apx:nth} shows a potential implementation of \texttt{nth} in Liquidity (\lstref{lst:nth_liq}) using a loop, and how the respective Michelson code looks like after compiling the contract with the Liquidity compiler (\lstref{lst:nth_tz}). \\
In case the \texttt{nth} is added as a predefined primitive to Michelson for the type \texttt{'a list}, its selection and typing rules are specified as follows:
\begin{lstlisting}[caption=Selection and type rule of the list \texttt{nth} instruction, language=, label=lst:nth_type]
:: (list 'a) : nat : 'A   ->  	option 'a : 'A

> NTH / l : index : S  =>  Some 'a: S
     iff index is within bounds
> NTH / l : index : S  =>  None 'a: S
     iff index is out of  bounds
\end{lstlisting}

As Michelson lists are represented with OCamls list type within the VM \cite{tezos_repo}, the computation for \texttt{nth} is simply applying the \texttt{nth} function of OCaml's List module, as shown in \lstref{lst:nth_impl}.
\begin{lstlisting}[caption=Simplified evaluation of \texttt{nth} in the Michelson interpreter, language=, label=lst:nth_impl]
match (instruction, stack) with
...
| (Nth ({elements = []; _}, (_, rest))) ->
  return ((None, rest)
| (Nth ({elements = es; length}, (index, rest))) ->
  let l_length = of_int length in
  if index <= l_length
    then return (Some (List.nth es index), rest)
    else return (None, rest)
\end{lstlisting}
\lstset{upquote=false}

\section{Type check}
\draft{}
\begin{itemize}
\item How are entry points assertions mapped to entrypoints?
\item Rules (e.g. overlapping entrypoint are not allowed, usage of tags)
\item Get entrypoint list \& types from chain if parent contract is already originated
\end{itemize}

>>>>

\section{Extensions to Michelson}
The way Michelson is interpreted is purely functional; it takes the current stack and an operation and builds a return stack from the initial one, without causing any side effects \cite{tezos_docs}. The recursive Michelson interpreter is defined as as a list of rules comprising of all possible inputs, i.e. program and stack types, and the respective output stack type of the computation if a rule applies. Each rule is of the following form: 
\begin{lstlisting}[caption=Rules form in the Michelson interpreter \cite{tezos_docs}, language=, numbers=none, label=lst:rules]
> (syntax pattern) / (initial stack pattern)  =>  (result stack pattern)
    iff (conditions)
    where (recursions)
    and (more recursions)
\end{lstlisting}
For each valid program and initial stack exactly one rule matches (given that any extra conditions over values on the stack, stated after the \texttt{iff} keyword are true) \cite{tezos_docs}. If the result depends on the results of other program interpretations (given after the keywords \texttt{iff, where, and} in rule form), the rule only applies if these partial results also match the respective intermediate result stack patterns.\\
In addition, there exists a typing rule for each syntax construct that restricts the valid input stacks. The typing rules use the meta variables \texttt{'a} for type and \texttt{'A} for stack type variables in order to express consistency within the program (not polymorphism). These rules are given in the following form:
\begin{lstlisting}[caption=Form of typing rules in Michelson's specification \cite{tezos_docs}, language=, numbers=none, label=lst:type_rules]
(syntax pattern)
:: (type of stack before) -> (type of stack after) [rule-name]
   iff (premises)
\end{lstlisting}
The type notations, as well as syntax and stack patterns are listed in \cite{tezos_docs}.\\

From the given formulas and assertion examples so far, one can derive a set of necessary instructions that need be be present in the target language. The following subsections list the instructions which have to be added to Michelson's instruction set, specify their selection and typing rules and describe how they can be implemented in the interpreter. These do not necessarily include high level instructions, like \texttt{sqrt} from \eqref{eq:prime}, that can be reproduced by using lower level instructions. For these purposes, a later iteration of the syntax will provide the feature of user-defined functions.

\subsection{Random}
A vital instruction that cannot be expressed with Michelson's current instruction set (state of protocol 008 Edo \cite{tezos_docs}) is the generation of a random value of certain types. Since smart contracts need to be deterministic, s.t. the network can reach a consensus about the state of the ledger \cite{chatterjee_probabilistic_2019}, the prevalent workarounds for generating pseudo-random numbers cannot be used for assertion checking. Distributed assertion checking explicitly requires as many validators as possible to generate a unique value, whereas common schemes for randomness, like oracles or using block attributes as seeds \cite{chatterjee_probabilistic_2019}, will provide the same values for all validators.\\
\lstref{lst:rand_type} specifies the selection and typing rule of a new instruction \texttt{rand}, which consumes an offset and a positive range from the stack and pushes a randomly generated integer to the top. This instruction, of course, can also be provided for other data types, such as \texttt{nat}, \texttt{string} or \texttt{mutez}.
\lstset{upquote=true}
\begin{lstlisting}[caption=Selection and typing rules of the integer \texttt{rand} instruction, language=, numbers=none, label=lst:rand_type]
:: int : nat: 'A -> int : 'A
> RAND/ offset : range : S  => int : S
\end{lstlisting}

Within the Michelson interpreter, the selection of rules is implemented as a huge match case \cite{tezos_repo}. If the rule for \texttt{rand} applies, the respective computation in OCaml could look as drafted in \lstref{lst:rand_impl}.
\begin{lstlisting}[caption=Simplified evaluation of \texttt{rand} in the Michelson interpreter, language=, label=lst:rand_impl]
match (instruction, stack) with
...
| (Rand (offset, (range, rest))) -> 
   Random.self_init;
   let rand_int = offset + Random.int(range) in
   return (rand_int, rest)  (* Resulting stack *)
\end{lstlisting}

\subsection{Nth}
An exception for not adding higher level instructions may be the more fundamental operation \texttt{nth} for accessing list elements, as it is used (sometimes more than once, ref. \eqref{eq:sorted_v2}) in many use cases. Adding an own predefined primitive for this reduces the origination costs of the assertion contract, given that its equivalent in the current instruction set requires the compiler to generate a workaround using iteration, for instance a loop or list iterator. As a first approximation, appendix \ref{apx:nth} shows a potential implementation of \texttt{nth} in Liquidity (\lstref{lst:nth_liq}) using a loop, and how the respective Michelson code looks like after compiling the contract with the Liquidity compiler (\lstref{lst:nth_tz}). \\
In case the \texttt{nth} is added as a predefined primitive to Michelson for the type \texttt{'a list}, its selection and typing rules are specified as follows:
\begin{lstlisting}[caption=Selection and type rule of the list \texttt{nth} instruction, language=, label=lst:nth_type]
:: (list 'a) : nat : 'A -> option 'a : 'A
> NTH / l : index : S  =>  Some 'a: S
     iff index is within bounds
> NTH / l : index : S  =>  None 'a: S
     iff index is out of  bounds
\end{lstlisting}
As Michelson lists are represented with OCamls list type within the VM \cite{tezos_repo}, the computation for \texttt{nth} could look as shown in \lstref{lst:nth_impl}.
\begin{lstlisting}[caption=Simplified evaluation of \texttt{nth} in the Michelson interpreter, language=, label=lst:nth_impl]
match (instruction, stack) with
...
| (Nth ({elements = []; _}, (_, rest))) ->
  return ((None, rest)
| (Nth ({elements = es; length}, (index, rest))) ->
  let l_length = of_int length in
  if index <= l_length
    then return (Some (List.nth es index), rest)
    else return (None, rest)
\end{lstlisting}
\lstset{upquote=false}

\section{Type check}
The task of the type checker is to map the assertions from the given file to the respective entrypoints of the parent contract. To this end, the parameter types have to be compared and matched against each other in order to find a correct assignment. The type checker either reads the code of the parent contract from a given file or retrieves the code from a given address on the blockchain. It must be ensured that the created mapping is injective and unambiguous, i.e., each assertion matches exactly one entrypoint and no entrypoint is covered by more than one assertion.\\
As explained in \secref{sec:tezos}, entrypoints can be selected by calling the default entrypoint and wrapping the parameter into the respective union constructors. Alternatively, they can be called explicitly using their tag and passing the parameter as raw value. The same applies for the assertions - by omitting the tag, the parameter type must be declared in respect to the default entrypoint type. If the assertion states an explicit tag, it may declare the raw parameter type. The tags do not necessarily have to be identical with the tags of the parent contract; in some cases, they may be chosen differently for documentation or for readability purposes. However, the tags can and should be used to resolve ambiguity in the mapping between assertions and entrypoints, that is if several entrypoints share the same input type. Ambiguity can also be caused by the fact that entrypoints can be sub-entrypoints of others, thus assertions overlap if a separate assertion is declared for both super- and sub-entrypoint. This is resolved by detecting overlapping assertions during the type check and rejecting assertion contracts where appropriate. \\
As an exemplification, consider a Michelson contract with five entrypoints (including the default entrypoint):
\begin{lstlisting}[numbers=none, language=Michelson]
parameter (or (int %A) (or %BC (int %B) (int %C)))
\end{lstlisting}
The following assertion contract for the given PC contains some valid and some invalid assertions:
\begin{lstlisting}[language=Assertion]
(assertion %A (i : int) ...)			 		  (* valid *)
(assertion %D (i : int) ...)			 		  (* invalid *)
(assertion (right (left (i : int)) ...)	(* valid *)
(assertion %BC (x : (or int int)) ...)	(* invalid *)
\end{lstlisting}
Due to its tag, the fist assertion can be assigned unambiguously to entrypoint A, whereas for the second it is not apparent whether it should be assigned to entrypoint B or C. Since the tag is omitted in the third assertion, the parameter type is given in respect to the default entrypoint and is assigned to B. The last assertion is invalid because it is overlapping with the previous assertion, which was already assigned to entrypoint B.
